---
title: "Lia_HSM"
output: html_document
date: "2026-04-06"
---
Attempting HSM
Starting from where we got in class after making Random Forest
Goal: ensemble with random forest and maxent

Libraries
```{r}
library(tidyverse)
library(raster)
library(terra)
library(randomForest)
library(rJava)
library(PresenceAbsence)
library(here)
library(ggplot2)
library(biomod2)
library(gstat)
```


Load final data from Git
```{r}
mosquito_final <- read_csv(here("Natalie/data_cleaned/mosquito_final_NDVI_MNDWI.csv"))
```


Preparing training data (from class 3/30, following Natalie's code), creating absences
```{r}
#add species columns
mosquito_final$cq <- NA
mosquito_final$ct <- NA
mosquito_final$aa <- NA
mosquito_final$count <- mosquito_final$Males + mosquito_final$Females

# create presence and absence for each species
mosquito_final$cq <- as.integer(mosquito_final$Species == "Culex quinquefasciatus" & mosquito_final$count > 0)
mosquito_final$ct <- as.integer(mosquito_final$Species == "Culex tarsalis" & mosquito_final$count > 0)
mosquito_final$aa <- as.integer(mosquito_final$Species == "Aedes aegypti" & mosquito_final$count > 0)

# merge rows (group_by) and summarize the three species
mosq_new <- mosquito_final |>
  group_by(
    Latitude, Longitude, YW, tavg1, tavg2, tmin1, tmin2, tmax1, tmax2, prec1, prec2, humd1, humd2, NDVI, MNDWI
  ) |>
  summarize(
    cq=max(cq), ct=max(ct), aa=max(aa)
  )
# omit NAs
mosq_new <- na.omit(mosq_new)

# determine presence and absence numbers (checking numbers)
sum(mosq_new$cq) #100918 observations
sum(mosq_new$ct) #36995 observations
sum(mosq_new$aa) #79614 observations

# Creating absent frames
cq0 <- subset(mosq_new, cq==0)
ct0 <- subset(mosq_new, ct==0)
aa0 <- subset(mosq_new, aa==0)

# creating present frame
cq1 <- subset(mosq_new, cq==1)
ct1 <- subset(mosq_new, ct==1)
aa1 <- subset(mosq_new, aa==1)

### equalize the number of presences and absences 
#**Switch for absent if need to make equal!**
cq0_samp <- cq0[sample(nrow(cq0),nrow(cq1), replace = FALSE),]
ct0_samp <- ct0[sample(nrow(ct0),nrow(ct1), replace = FALSE),]
aa0_samp <- aa0[sample(nrow(aa0),nrow(aa1), replace = FALSE),]

# now bind the absences and presences together
cq <- rbind(cq1, cq0_samp)
ct <- rbind(ct1, ct0_samp)
aa <- rbind(aa1, aa0_samp)

#Natalie saved csv, but I'll just use hers since she did that already
#write.csv(cq, here("Natalie/data_cleaned/cq_df.csv"))
#write.csv(ct, here("Natalie/data_cleaned/ct_df.csv"))
#write.csv(aa, here("Natalie/data_cleaned/aa_df.csv")) 
```


Train & test model
```{r}
#read in the data we just made
cq <- read.csv(here("Natalie/data_cleaned/cq_df.csv"))
ct <- read.csv(here("Natalie/data_cleaned/ct_df.csv"))
aa<- read.csv(here("Natalie/data_cleaned/aa_df.csv"))

#cq train model, test to ensure no overfitting
train_indices_cq <- sample(1:nrow(cq), size = 0.8 *nrow(cq)) #taking a random sample from cq
train_data_cq <- cq[train_indices_cq,]
test_data_cq <- cq[-train_indices_cq,]
nrow(train_data_cq) #161468
nrow(test_data_cq) #40368

#write.csv(train_data, file = here("Natalie/data_cleaned/test_train/train_data_cq.csv"))
#write.csv(test_data, file = here("Natalie/data_cleaned/test_train/test_data_cq.csv"))

#ct train model
train_indices_ct <- sample(1:nrow(ct), size = 0.8 *nrow(ct)) #taking a random sample from ct
train_data_ct <- ct[train_indices_ct,]
test_data_ct <- ct[-train_indices_ct,]

#aa train model
train_indices_aa <- sample(1:nrow(aa), size = 0.8 *nrow(aa)) #taking a random sample from cq
train_data_aa <- cq[train_indices_aa,]
test_data_aa <- cq[-train_indices_aa,]

```


GLMs

cq
```{r}
###cq first
#cq 2 weeks with water
glm_model_cq2_h20 <- glm(cq ~ NDVI + MNDWI + tavg2+ tmin2+ tmax2 + prec2 + humd2, 
                         data = train_data_cq, 
                         family = binomial(link = "logit"))

#cq 1 week with water
glm_model_cq1_h20 <- glm(cq ~ NDVI + MNDWI + tavg1 + tmin1 + tmax1 + prec1 + humd1, 
                         data = train_data_cq, 
                         family = binomial(link = "logit"))

#cq 2 weeks without water
glm_model_cq2 <- glm(cq ~ tavg2 + tmin2+ tmax2 + prec2 + humd2, 
                         data = train_data_cq, 
                         family = binomial(link = "logit"))

#cq 1 week without water
glm_model_cq1 <- glm(cq ~ tavg1 + tmin1+ tmax1 + prec1 + humd1, 
                         data = train_data_cq, 
                         family = binomial(link = "logit"))

```
ct
```{r}
###ct 
#ct 2 weeks with water
glm_model_ct2_h20 <- glm(ct ~ NDVI + MNDWI + tavg2+ tmin2+ tmax2 + prec2 + humd2, 
                         data = train_data_ct, 
                         family = binomial(link = "logit"))

#ct 1 week with water
glm_model_ct1_h20 <- glm(ct ~ NDVI + MNDWI + tavg1 + tmin1 + tmax1 + prec1 + humd1, 
                         data = train_data_ct, 
                         family = binomial(link = "logit"))

#ct 2 weeks without water
glm_model_ct2 <- glm(ct ~ tavg2 + tmin2+ tmax2 + prec2 + humd2, 
                         data = train_data_ct, 
                         family = binomial(link = "logit"))

#ct 1 week without water
glm_model_ct1 <- glm(ct ~ tavg1 + tmin1+ tmax1 + prec1 + humd1, 
                         data = train_data_ct, 
                         family = binomial(link = "logit"))
```
aa
```{r}
###aa
#aa 2 weeks with water
glm_model_aa2_h20 <- glm(aa ~ NDVI + MNDWI + tavg2+ tmin2+ tmax2 + prec2 + humd2, 
                         data = train_data_aa, 
                         family = binomial(link = "logit"))

#aa 1 week with water
glm_model_aa1_h20 <- glm(aa ~ NDVI + MNDWI + tavg1 + tmin1 + tmax1 + prec1 + humd1, 
                         data = train_data_aa, 
                         family = binomial(link = "logit"))

#aa 2 weeks without water
glm_model_aa2 <- glm(aa ~ tavg2 + tmin2+ tmax2 + prec2 + humd2, 
                         data = train_data_aa, 
                         family = binomial(link = "logit"))

#aa 1 week without water
glm_model_aa1 <- glm(aa ~ tavg1 + tmin1+ tmax1 + prec1 + humd1, 
                         data = train_data_aa, 
                         family = binomial(link = "logit"))
```



Random forest model

cq
```{r}
#From Desiree
#cq 2 weeks w water
rf_model_cq2_h2o <- randomForest(x = train_data_cq[, c("tavg2", "tmin2", "tmax2", "prec2", "humd2", "NDVI", "MNDWI")], 
                                 y=as.factor(train_data_cq$cq), 
                                 nodesize = 10, 
                                 ntree = 100)

#cq 1 week w water
rf_model_cq1_h2o <- randomForest(x = train_data_cq[, c("tavg1", "tmin1", "tmax1", "prec1", "humd1", "NDVI", "MNDWI")], 
                                 y=as.factor(train_data_cq$cq), 
                                 nodesize = 10, 
                                 ntree = 100)

#cq 2 weeks no water
rf_model_cq2 <- randomForest(x = train_data_cq[, c("tavg2", "tmin2", "tmax2", "prec2", "humd2")], 
                                 y=as.factor(train_data_cq$cq), 
                                 nodesize = 10, 
                                 ntree = 100)

#cq 1 week no water
rf_model_cq1 <- randomForest(x = train_data_cq[, c("tavg1", "tmin1", "tmax1", "prec1", "humd1")], 
                                 y=as.factor(train_data_cq$cq), 
                                 nodesize = 10, 
                                 ntree = 100)
```
ct
```{r}
#ct 2 weeks w water
rf_model_ct2_h2o <- randomForest(x = train_data_ct[, c("tavg2", "tmin2", "tmax2", "prec2", "humd2", "NDVI", "MNDWI")], 
                                 y=as.factor(train_data_ct$ct), 
                                 nodesize = 10, 
                                 ntree = 100)

#ct 1 week w water
rf_model_ct1_h2o <- randomForest(x = train_data_ct[, c("tavg1", "tmin1", "tmax1", "prec1", "humd1", "NDVI", "MNDWI")], 
                                 y=as.factor(train_data_ct$ct), 
                                 nodesize = 10, 
                                 ntree = 100)

#ct 2 weeks no water
rf_model_ct2 <- randomForest(x = train_data_ct[, c("tavg2", "tmin2", "tmax2", "prec2", "humd2")], 
                                 y=as.factor(train_data_ct$ct), 
                                 nodesize = 10, 
                                 ntree = 100)

#ct 1 week no water
rf_model_ct1 <- randomForest(x = train_data_ct[, c("tavg1", "tmin1", "tmax1", "prec1", "humd1")], 
                                 y=as.factor(train_data_ct$ct), 
                                 nodesize = 10, 
                                 ntree = 100)
```
aa
```{r}
#aa 2 weeks w water
rf_model_aa2_h2o <- randomForest(x = train_data_aa[, c("tavg2", "tmin2", "tmax2", "prec2", "humd2", "NDVI", "MNDWI")], 
                                 y=as.factor(train_data_aa$aa), 
                                 nodesize = 10, 
                                 ntree = 100)

#aa 1 week w water
rf_model_aa1_h2o <- randomForest(x = train_data_aa[, c("tavg1", "tmin1", "tmax1", "prec1", "humd1", "NDVI", "MNDWI")], 
                                 y=as.factor(train_data_aa$aa), 
                                 nodesize = 10, 
                                 ntree = 100)

#aa 2 weeks no water
rf_model_aa2 <- randomForest(x = train_data_aa[, c("tavg2", "tmin2", "tmax2", "prec2", "humd2")], 
                                 y=as.factor(train_data_aa$aa), 
                                 nodesize = 10, 
                                 ntree = 100)

#aa 1 week no water
rf_model_aa1 <- randomForest(x = train_data_aa[, c("tavg1", "tmin1", "tmax1", "prec1", "humd1")], 
                                 y=as.factor(train_data_aa$aa), 
                                 nodesize = 10, 
                                 ntree = 100)
```



Maximum Entropy model (following the book)
*book says not to use maxent if we have both presence and absence data*
what is a true absence?
```{r}
maxent_path <- system.file("java", package="dismo")
maxent_path


```

AUC results
```{r}
#Calculating AUC from GLMs
glm_auc_df <- data.frame(rf_test_pred[,1], test_presence, glm_test_pred[,2])
auc_glm <- auc(glm_auc_df, which.model =1, na.rm = TRUE)
  
#Calculating AUC from RFs
rf_auc_df <- data.frame(rf_test_pred[,1], test_presence, rf_test_pred[,2])
auc_rf <- auc(rf_auc_df, which.model =1, na.rm = TRUE)
```


Ensemble
```{r}

```

-------
Notes from class 4/6/2026:

ecological niche factor analysis (in envelopes and distance based approaches in book part III chapter 9, page 161)
```{r}
#GLM cq predict using predict function
#to visualize whole cq dataset, input into new column in cq
cq$glm_cq2_h2o_predict <- predict(glm_model_cq2_h20, cq) #why are some negative

#plotting
library(leaflet)
cq_map_2_h2o <- leaflet() |> 
  addProviderTiles(providers$Esri) |> 
  addCircleMarkers(data = cq, ~Longitude, ~Latitude, radius = 1, color = cq$glm_cq2_h2o_predict)

cq_map_2_h2o

#making a subset with less data to visualize using one week
test <- cq |> filter(
  YW == "YW202128"
)

#making a map with ggplot

#glm_model_cq2_h20 <- read_rds(from box)
#prec2 <- raster(path to yw202128 tif)
#copy for all env vars #is this on box? 
#identify a range of weeks with the most activity to do this for #use tidy to group by week and summarize
#then read all the rasters into a stack
#env202128 <- stack(NDVI, MNDWI, tavg2, tmin2, tmax2, prec2, humd2)
#plot(env202128) #to see all layers
#create a raster layer of the prediction
#cq202128 <- predict(env202128, glm_model_cq2_h2o)

#use true skill statistic to look at true positive and negative rates of training data on test data, determines a threshold of deciding presence or absence

```

Inspiration from Desiree on how to present data:
sensitivity, specificity, maxSS
response curves to env vars
identify where we expect to see species in a given week or over time, or hotspots
categories: maps, response curves, accuracy/model fit numbers of each model, variable significance, comparing the fit of dif models, example or aggregated output

creating response curves
```{r}
cq_sub <- mosquito_final[sample(nrow(mosquito_final), 10000),]

library(biomod2)
library(ggplot2)

rp <- response.plot2(model = c("glm_model_cq2_h2o"), 
                     data = cq_sub[, c("NDVI", "MNDWI", "tavg", "tmin2", "tmax2", "prec2", "humd2")], 
                     show.variables = c("NDVI", "MNDWI", "tavg", "tmin2", "tmax2", "prec2", "humd2"), 
                     fixed.var.metric = "mean", 
                     plot = FALSE, use.formal.names = TRUE) 
#response plot, we will use multiple models in future
#response.plot2 is depreciated, doesn't work

gg.rp <- rp |> ggplot(aes = expl.val, y = pred.val)
```
*pick weeks for env tifs*
*pick 1 or 2 week lag for all env vars*
*use bm_PlotResponseCurves*





