---
title: "Desiree_Cide"
output: html_document
date: "2026-03-24"
---
load library
```{r}
library(tidyverse)
```

create columns
```{r}
mosquito$cq <- NA
mosquito$ct <- NA
mosquit$aa <- NA
mosquito$count <- mosquito$male + mosquito$female



```


Change count to interger
```{r}
mosquito$cq <- as.integer(mosquito$Species == "Culex quinquefasciatus" & mosquito$count>0)
mosquito$ct <- as.integer(mosquito$Species == "Culex tarsalis" & mosquito$count>0)
mosquito$aa <- as.integer(mosquito$Species == "Aedes aegypti" & mosquito$count>0)


```



Merge rows Makes new data frame, decreases the observations by a good chunk
```{r}

mosq_new <- mosquito %>% group_by(latitude, longitude, yw, CLIMATE) #need to add all the climate variables
summarise(cq=max(cq), ct=max(ct), aa=max(aa))


```

Na omit 
```{r}
mosq_new <-na.omit(mosq_new) #desiree amount is ~38000
```


determine number of present / absent
```{r}
sum(mosq_new$cq) #gives number of observations where the species is present
sum(mosq_new$ct)#gives number of observations where the species is present
sum(mosq_new$aa)#gives number of observations where the species is present
```

Creating absent for species
```{r}
cq0 <- subset(mosq_new, cq==0)
ct0 <- subset(mosq_new, ct==0)
aa0 <- subset(mosq_new, aa==0)
```

create present for species
```{r}
cq1 <- subset(mosq_new, cq==1)
ct1 <- subset(mosq_new, ct==1)
aa1 <- subset(mosq_new, aa==1)
```

Sampling the data frame to get equal data frame amounts for present and absent
**switch if more present, need the amount of to be equal**
```{r}
cq0_samp <- cq0[sample(nrow(cq0), nrow(cq1), replace = FALSE),]
ct0_samp <- ct0[sample(nrow(ct0), nrow(ct1), replace = FALSE),]
aa0_samp <- aa0[sample(nrow(aa0), nrow(aa1), replace = FALSE),]


```


Putting it all together, all the data frame together, create a balance happy model
```{r}
cq <- rbind(cq1, cq0_samp)
ct <- rbind(ct1, ct0_samp)
aa <- rbind(aa1, aa0_samp)

```





