---
title: "SDM_Calculation"
output: html_document
date: "2026-03-01"
---

Loading Packages
```{r}
library(tidyverse) #data management
library(dplyr) # data management
library(here) # to use here directory to open data files easier
library(raster)
library(terra)
library(randomForest)
library(rJava)
library(ggplot2)
library(biomod2)
library(gstat)
library(leaflet)

library(adehabitatHS) #PCA, page 161, ENFA
library(pROC)


```


Loading in Data
```{r}

cq <- read.csv(here("Natalie/data_cleaned/cq_df.csv"))
ct <- read.csv(here("Natalie/data_cleaned/ct_df.csv"))
cq<- read.csv(here("Natalie/data_cleaned/cq_df.csv"))

```

Making Spatial Rasters for environment data  **Testing to make easier for Maxent?**
```{r}

cq_vec <- terra :: vect(cq, geom = c("Latitude", "Longitude"), crs = "EPSG:4326") #making a spat_raster
plot(cq_vec)

template <- terra :: rast( xmin =33.2043, xmax= 33.7987, ymin= -112.573997704, ymax = -111.5837, res = 0.01, crs = "EPSG:4326")



```


Test and train Data frames
```{r}
#splitting into training data --> cq
train_indices_cq <- sample(1:nrow(cq), size = 0.8 *nrow(cq)) #taking a random sample from cq
train_data_cq <- cq[train_indices,]
test_data_cq <- cq[-train_indices,]
nrow(train_data_cq) 
nrow(test_data_cq) 

write.csv(train_data, file = here("Natalie/data_cleaned/test_train/train_data_cq.csv"))
write.csv(test_data, file = here("Natalie/data_cleaned/test_train/test_data_cq.csv"))



# Ct
train_indices_ct <- sample(1:nrow(ct), size = 0.8 *nrow(ct)) #taking a random sample from cq
train_data_ct <- ct[train_indices,]
test_data_ct <- ct[-train_indices,]
nrow(train_data_ct) 
nrow(test_data_ct) 

write.csv(train_data_ct, file = here("Natalie/data_cleaned/test_train/train_data_ct.csv"))
write.csv(test_data_ct, file = here("Natalie/data_cleaned/test_train/test_data_ct.csv"))



#cq
train_indices_cq <- sample(1:nrow(cq), size = 0.8 *nrow(ct)) #taking a random sample from cq
train_data_ct <- ct[train_indices,]
test_data_ct <- ct[-train_indices,]
nrow(train_data_ct) 
nrow(test_data_ct) 

write.csv(train_data_ct, file = here("Natalie/data_cleaned/test_train/train_data_ct.csv"))
write.csv(test_data_ct, file = here("Natalie/data_cleaned/test_train/test_data_ct.csv"))
```


GLM - cq
```{r}
glm_model_cq2_h20 <- glm(cq ~ NDVI+ MNDWI+
                                 tavg2+ tmin2+ tmax2+
                                 prec2+ humd2, data = train_data, family = binomial(link = "logit"))


#saveRDS(glm_model_cq2_h20, file = "glm_model_cq2_h20.rds")







glm_model_cq1_h2o <- glm(cq ~ NDVI+ MNDWI+
                                 tavg1+ tmin1+ tmax1+
                                 prec1+ humd1, data = train_data, family = binomial(link = "logit"))


glm_model_cq2 <- glm(cq ~ tavg2+ tmin2+ tmax2+
                                 prec2+ humd2, data = train_data, family = binomial(link = "logit"))


glm_model_cq1 <- glm(cq ~ tavg1+ tmin1+ tmax1+
                                 prec1+ humd1, data = train_data, family = binomial(link = "logit"))


```


GLM Cq Predict
```{r}
#glm_model_cq2_h20

cq$glm_cq2_h2o_predict <- predict(glm_model_cq2_h20, cq) #negative?


cq_map_2H2o <- leaflet() %>% 
  addProviderTiles(providers$Esri)%>%
    addCircleMarkers(data = test, ~Longitude, ~Latitude, radius = 1, color= cq$glm_cq2_h2o_predict)

test <- cq %>% filter(
  YW == "YW202128"
)


cq_map_2H2o
  
  
mosquito_map <- leaflet() %>% 
  addProviderTiles(providers$Esri) %>% 
  addCircleMarkers(data = mosquito_reduced, ~Longitude, ~Latitude, radius = 1)



```




GLM ct

```{r}
glm_model_ct2_h20 <- glm(ct ~ NDVI+ MNDWI+
                                 tavg2+ tmin2+ tmax2+
                                 prec2+ humd2, data = train_data, family = binomial(link = "logit"))

glm_model_ct1_h2o <- glm(ct ~ NDVI+ MNDWI+
                                 tavg1+ tmin1+ tmax1+
                                 prec1+ humd1, data = train_data, family = binomial(link = "logit"))

glm_model_ct2 <- glm(ct ~ tavg2+ tmin2+ tmax2+
                                 prec2+ humd2, data = train_data, family = binomial(link = "logit"))


glm_model_ct1 <- glm(ct ~ tavg1+ tmin1+ tmax1+
                                 prec1+ humd1, data = train_data, family = binomial(link = "logit"))


```


GLM cq **aa**
```{r}
glm_model_cq2_h20 <- glm(cq ~ NDVI+ MNDWI+
                                 tavg2+ tmin2+ tmax2+
                                 prec2+ humd2, data = train_data, family = binomial(link = "logit"))

glm_model_cq1_h2o <- glm(cq ~ NDVI+ MNDWI+
                                 tavg1+ tmin1+ tmax1+
                                 prec1+ humd1, data = train_data, family = binomial(link = "logit"))

glm_model_cq2 <- glm(cq ~ tavg2+ tmin2+ tmax2+
                                 prec2+ humd2, data = train_data, family = binomial(link = "logit"))


glm_model_cq1 <- glm(cq ~ tavg1+ tmin1+ tmax1+
                                 prec1+ humd1, data = train_data, family = binomial(link = "logit"))


```



RF-cq
```{r}

rf_model_cq1_h2o <- randomForest(x = train_data[, c("NDVI","MNDWI",
                       "tavg1","tmin1", "tmax1", "prec1", "humd1")], y=as.factor(train_data$cq), nodesize = 10, ntree = 100)

varImpPlot(rf_model_cq1_h2o)

rf_model_cq2_h2o <- randomForest(x = train_data[, c("NDVI","MNDWI",
                      "tavg2","tmin2", "tmax2", "prec2", "humd2")], y=as.factor(train_data$cq), nodesize = 10, ntree = 100)
varImpPlot(rf_model_cq2_h2o)

rf_model_cq1 <- randomForest(x = train_data[, c("tavg1","tmin1", "tmax1", "prec1", "humd1")], y=as.factor(train_data$cq), nodesize = 10, ntree = 100)
varImpPlot(rf_model_cq1)

rf_model_cq2 <- randomForest(x = train_data[, c("tavg2","tmin2", "tmax2", "prec2", "humd2")], y=as.factor(train_data$cq), nodesize = 10, ntree = 100)
varImpPlot(rf_model_cq2)


```


RF -ct
```{r}
rf_model_ct1_h2o <- randomForest(x = train_data[, c("NDVI","MNDWI",
                       "tavg1","tmin1", "tmax1", "prec1", "humd1")], y=as.factor(train_data$ct), nodesize = 10, ntree = 100)

varImpPlot(rf_model_ct1_h2o)

rf_model_ct2_h2o <- randomForest(x = train_data[, c("NDVI","MNDWI",
                      "tavg2","tmin2", "tmax2", "prec2", "humd2")], y=as.factor(train_data$ct), nodesize = 10, ntree = 100)
varImpPlot(rf_model_ct2_h2o)

rf_model_ct1 <- randomForest(x = train_data[, c("tavg1","tmin1", "tmax1", "prec1", "humd1")], y=as.factor(train_data$ct), nodesize = 10, ntree = 100)
varImpPlot(rf_model_ct1)

rf_model_ct2 <- randomForest(x = train_data[, c("tavg2","tmin2", "tmax2", "prec2", "humd2")], y=as.factor(train_data$ct), nodesize = 10, ntree = 100)
varImpPlot(rf_model_ct2)


```


rf - cq
```{r}
rf_model_cq1_h2o <- randomForest(x = train_data[, c("NDVI","MNDWI",
                       "tavg1","tmin1", "tmax1", "prec1", "humd1")], y=as.factor(train_data$cq), nodesize = 10, ntree = 100)

varImpPlot(rf_model_cq1_h2o)

rf_model_cq2_h2o <- randomForest(x = train_data[, c("NDVI","MNDWI",
                      "tavg2","tmin2", "tmax2", "prec2", "humd2")], y=as.factor(train_data$cq), nodesize = 10, ntree = 100)
varImpPlot(rf_model_cq2_h2o)

rf_model_cq1 <- randomForest(x = train_data[, c("tavg1","tmin1", "tmax1", "prec1", "humd1")], y=as.factor(train_data$cq), nodesize = 10, ntree = 100)
varImpPlot(rf_model_cq1)

rf_model_cq2 <- randomForest(x = train_data[, c("tavg2","tmin2", "tmax2", "prec2", "humd2")], y=as.factor(train_data$cq), nodesize = 10, ntree = 100)
varImpPlot(rf_model_cq2)


```


Calculating AUC **NEED PREDCIT?? help me :(**
```{r}
#Creating Data frame of observed and predicted values

#extracting test data and coords to test models.
test_coords_cq <- test_data_cq[, c("Longitude", "Latitude")]
test_presence_cq <- test_data_cq$cq


#glm
glm_test_pred <-extract(glm_predict, test_coords, pos_label = 1)
```



Guisan AUC Previous work, RF and Maxent,  
```{r}
#random forest#random forglm_predictest
rf_test_pred <- extract(rf_pred, test_coords, pos_label = 1) 

#Calculating AUC

glm_auc_df <- data.frame(rf_test_pred[,1], test_presence, glm_test_pred[,2])
auc_glm <- auc(glm_auc_df, which.model =1, na.rm = TRUE)
  
rf_auc_df <- data.frame(rf_test_pred[,1], test_presence, rf_test_pred[,2])
auc_rf <- auc(rf_auc_df, which.model =1, na.rm = TRUE)

maxent_auc_df <- data.frame(rf_test_pred[,1], test_presence, maxent_test_pred)


auc_results <- rbind(auc_glm, auc_rf, auc_maxent)
colnames(auc_results) <- c("AUC", "StdDev")
rownames(auc_results) <-c("GLM", "Rf", "Maxent")


```


Random Forest Model **Code from previous SDM calc**
```{r}
#data prep

train_data$presence_factor <- as.factor(train_data$presence) #changes the data into factors 

rf_model <- randomForest(presence_factor ~ Mean_Temp_of_Wettest_Quarter + Isothermality + Mean_Temp_of_Wettest_Quarter + Mean_temp_of_Driest_quarter + Precipitation_of_wettest_Month + Precipitation_of_driest_month + Precipitation_Seasonality + Precipitation_of_Warmest_Quarter + Precipitation_of_coldest_Quart, data = train_data, ntree = 1000, importance = TRUE, na.action = na.omit) #thinks that there are na present, but there is none? confused. adding na.action = na.omit fixes the problem though

print(rf_model)

#variable importance
print(importance(rf_model)) #need to figure out what this means exactly number wise
varImpPlot(rf_model, main = "Random Forest: Bio Variable Importance")

#predict across study area
rf_pred <- predict(env_vars, rf_model, type = "prob", index = 2) #WHAT DOES TYPE AND INDEX MEAN?
#rf_test <- predict(rf_model_test, type = "prob", index =2 ) # how it is done in book, 

plot(rf_pred, main = "RF Model Probability: Amsinkia intermedia", xlab = "lon", ylab = "lat") 
points(lon_lat, pch =10, col = "red", cex =0.5)


save(rf_pred, file =here("Data_Analysis/Random_Forest/rf_pred_sp45.Rdata"))

```



Maxent Model **Code from previous work**
```{r}
maxent_path <- system.file("java", package="dismo")
maxent_path

#if file is telling maxent where the files are located

lon_lat_maxent <- presence_df[c("lon", "lat")]

if (file.exists(paste0(maxent_path, "/maxent.jar"))) {
  
prescence_matrix <- as.data.frame(lon_lat_maxent)

#convert spat raster to raster stack
env_vars_raster <- stack(env_vars)

#building maxent model
maxent_model <- maxent(env_vars_raster, prescence_matrix)
print(maxent_model)

#Predict the distribution

maxent_pred <- predict(maxent_model, env_vars_raster)

#Visuzalize prediction
plot(maxent_pred, main = "Maxent: Habitat Suitability - Amsinckia intermedia", xlab = "lon", ylab = "lat",)
 
#response curves
response(maxent_model
         )

} else {
  cat("MaxEnt not available (requires maxent.jar)\n")
  cat("To use MaxEnt:\n")
  cat("1. Download maxent.jar from: https://biodiversityinformatics.amnh.org/open_source/maxent/\n")
  cat("2. Place it in:", maxent_path, "\n")
  cat("3. Restart R\n\n")
  cat("Skipping MaxEnt for this exercise...\n")
  maxent_pred <- NULL}


save(maxent_pred, file =here("Data_Analysis/Maxent/maxent_pred_sp45.Rdata"))

```



Calculating AUC - Guisan Book Example
```{r}
#Creating Data frame of observed and predicted values

#extracting test data and coords to test models.
test_coords <- test_data[, c("lon", "lat")]
test_presence <- test_data$presence

#glm
glm_test_pred <-extract(glm_predict, test_coords, pos_label = 1)


#random forest#random forglm_predictest
rf_test_pred <- extract(rf_pred, test_coords, pos_label = 1) 

#Maxent 
 if (!is.null(maxent_pred)) {
  maxent_test_pred <- extract(maxent_pred, test_coords)
} else {
  maxent_test_pred <- NULL
}

#Calculating AUC

glm_auc_df <- data.frame(rf_test_pred[,1], test_presence, glm_test_pred[,2])
auc_glm <- auc(glm_auc_df, which.model =1, na.rm = TRUE)
  
rf_auc_df <- data.frame(rf_test_pred[,1], test_presence, rf_test_pred[,2])
auc_rf <- auc(rf_auc_df, which.model =1, na.rm = TRUE)

maxent_auc_df <- data.frame(rf_test_pred[,1], test_presence, maxent_test_pred)
auc_maxent <- auc(maxent_auc_df, which.model =1, na.rm = TRUE)


auc_results <- rbind(auc_glm, auc_rf, auc_maxent)
colnames(auc_results) <- c("AUC", "StdDev")
rownames(auc_results) <-c("GLM", "Rf", "Maxent")
```


Ensemble Model **code from previous work**
```{r}
#turning maxent to different spat for enseble purpose
maxent_pred_spat <- rast(maxent_pred)


#ensemble stacking, mean
pred_stack <- stack( x = c(glm_predict, rf_pred, maxent_pred_spat))

#ensemeble stacking, weighted means
weights <- auc_results$AUC / sum(auc_results$AUC)

ensemble_pred <- weighted.mean(pred_stack, w = weights)
plot(ensemble_pred)


#using test data now to get AUC
ensemble_test_pred <- extract(ensemble_pred, test_coords, pos_label = 1)


ensemble_auc_df <- data.frame(rf_test_pred[,1], test_presence, ensemble_test_pred)

auc_ensemble<- auc(ensemble_auc_df, which.model = 1, na.rm = TRUE)

auc_results <- rbind(auc_glm, auc_rf, auc_maxent, auc_ensemble)
colnames(auc_results) <- c("AUC", "StdDev")
rownames(auc_results) <-c("GLM", "Rf", "Maxent", "Ensemble")

write.csv(auc_results, file = here("data/02_clean/auc/auc_results.sp45"))
```




