---
title: "NJD_Mosquito_HSM"
author: "Nick Dabagia"
date: "2026-02-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# functions
```{r}
library(tidyverse)
library(lubridate)
library(randomForest)   # Random Forest algorithm
library(dismo)          # Core SDM functions
library(raster)         # Spatial data handling
library(sp)             # Spatial data classes
```
# read in data
```{r eval=FALSE, include=FALSE}
setwd("/Users/ndabagia/Desktop/classes//GlobalChange/HSM_mosquito")
mosquito_data <- read.csv("mosquito_final.csv", row.names=1)

```


Hmmmm...this picture is incomplete with the NAs

```{r eval=FALSE, include=FALSE}
#Use na.rm = TRUE to filter them out
mosquito_data %>% 
  group_by(Species) %>% 
  summarize(Males = sum(Males, na.rm = TRUE), Females = sum(Females, na.rm = TRUE))

# FILTER OUT ALL MOSQS THAT AREN'T CULEX OR AEDES

species_int <- mosquito_data |>
  filter(Species == "Aedes aegypti" | 
           Species == "Culex tarsalis"|
           Species == "Culex quinquefasciatus"|
           Species == "None"|
           Species == "No mosquitoes collected"|
           is.na(Species))

# add a week column to 1-52 mosquito dataset
species_int$Week <- strftime(species_int$Date, format = "%V")
species_int$Year <- strftime(species_int$Date, format = "%G" )

species_int <- species_int |>
  #mutate(
   # Date = as.Date(Date),
    #Week = isoweek(Date)
  #) %>%
  relocate(Week, .after = 1)|>
  relocate(Year, .after = 2)

# make a bounding box to filter the observations to bounded box for Maricopa county/metro phoenix
# West: -112.573997704 North: 33.7987N East: -111.5837 South: 33.2043
maricopa_obs <- species_int |>
  filter(Latitude >= 33.2043 & Latitude <= 33.7987,
         Longitude <= -111.5837 & Longitude  >=  -112.573997704
)

# sum(is.na(maricopa_obs$Females)) ensure zero NA's in key column

# add presence column a la Oskar
maricopa_obs %>% 
  mutate(Presence = ifelse(Males == 0 & Females == 0, 0, 1) )

# str(maricopa_obs)
maricopa_obs |>
  filter(Week > 52)|>
  print()

```
### write the resulting file to local
```{r}
setwd("/Users/ndabagia/Desktop/classes//GlobalChange/HSM_mosquito")

mosquito_final <- read.csv("mosquito_final.csv")

```

```{r eval=FALSE, include=FALSE}
setwd("/Users/ndabagia/Desktop/classes//GlobalChange/HSM_mosquito")
NDVI <- raster("NDVI.tif")
MNDWI <- raster("MNDWI.tif")

lonlat_df <- data.frame(
  Longitude = mosquito_final$Longitude,
  Latitude = mosquito_final$Latitude
)

mosquito_final$NDVI <- raster::extract(NDVI, lonlat_df)
mosquito_final$MNDWI <- raster::extract(MNDWI, lonlat_df)

plot(x = mosquito_final$Longitude, y = mosquito_final$Latitude, data = mosquito_final$NDVI)
plot(x = mosquito_final$Longitude, y = mosquito_final$Latitude, data = mosquito_final$MNDWI)
#these are crappy maps, but it shows some idea of where the sampling points are
#note to self to make better ones later in leaflet
# add a column for overall presence / absence
mosquito_final <- mosquito_final %>% 
mutate(
  Presence = ifelse(Males ==0 & Females == 0, 0, 1)
)
#write.csv(mosquito_final, "mosquito_final.csv")
```
## do the fancy manipulation to the data that Desiree recommended
```{r}
# create columns with each of the three species

mosquito_final$cq <- NA
mosquito_final$ct <- NA
mosquito_final$aa <- NA
mosquito_final$count <- mosquito_final$Males + mosquito_final$Females

# create presence and absence for each species
mosquito_final$cq <- as.integer(mosquito_final$Species == "Culex quinquefasciatus" & mosquito_final$count > 0)
mosquito_final$ct <- as.integer(mosquito_final$Species == "Culex tarsalis" & mosquito_final$count > 0)
mosquito_final$aa <- as.integer(mosquito_final$Species == "Aedes aegypti" & mosquito_final$count > 0)

# merge rows (group_by) and summarize the three species
mosq_new <- mosquito_final |>
  group_by(
    Latitude, Longitude, YW, tavg1, tavg2, tmin1, tmin2, tmax1, tmax2, prec1, prec2, humd1, humd2, NDVI, MNDWI
  ) |>
  summarize(
    cq=max(cq), ct=max(ct), aa=max(aa)
  )
# omit NAs
mosq_new <- na.omit(mosq_new)

# determine presence and absence numbers

sum(mosq_new$cq)
sum(mosq_new$ct)
sum(mosq_new$aa)



# subsetting to get all the absences

cq0 <- subset(mosq_new, cq==0)
ct0 <- subset(mosq_new, ct==0)
aa0 <- subset(mosq_new, aa==0)
# subsetting to get the presences
cq1 <- subset(mosq_new, cq==1)
ct1 <- subset(mosq_new, ct==1)
aa1 <- subset(mosq_new, aa==1)

### equalize the number of presences and absences 

cq0_samp <- cq0[sample(nrow(cq0),nrow(cq1), replace = FALSE),]
ct0_samp <- ct0[sample(nrow(ct0),nrow(ct1), replace = FALSE),]
aa0_samp <- aa0[sample(nrow(aa0),nrow(aa1), replace = FALSE),]

# now bind the absences and presences together

cq <- rbind(cq1, cq0_samp)
ct <- rbind(ct1, ct0_samp)
aa <- rbind(aa1, aa0_samp)

### 

```

# randomForest via Desiree for each mosquito species
```{r}
set.seed(123)
ct_rf <- randomForest(x=ct[, 4:15], y=as.factor(ct$ct), nodesize = 10, # xx[,4:15] accounts for the 11 predictors
                      ntree = 100) 
# nodesize helps with speed and avoids overfitting | ntree helps again with speed

cq_rf <- randomForest(x=cq[, 4:15], y=as.factor(cq$cq), nodesize = 10, 
                      ntree = 100) 
# nodesize helps with speed and avoids overfitting | ntree helps again with speed
aa_rf <- randomForest(x=aa[, 4:15], y=as.factor(aa$aa), nodesize = 10, 
                      ntree = 100) 
#_______________________________#
### take out ndvi and mndwi
ct_rf_simple <- randomForest(x=ct[, 4:13], y=as.factor(ct$ct), nodesize = 10, # xx[,4:15] accounts for the 11 predictors
                      ntree = 100) 
# nodesize helps with speed and avoids overfitting | ntree helps again with speed

cq_rf_simple <- randomForest(x=cq[, 4:13], y=as.factor(cq$cq), nodesize = 10, 
                      ntree = 100) 
# nodesize helps with speed and avoids overfitting | ntree helps again with speed
aa_rf_simple <- randomForest(x=aa[, 4:13], y=as.factor(aa$aa), nodesize = 10, 
                      ntree = 100) 

```
### variable importance
```{r}
randomForest:: varImpPlot(ct_rf)
randomForest:: varImpPlot(cq_rf)
randomForest:: varImpPlot(aa_rf)
randomForest:: varImpPlot(ct_rf_simple)
randomForest:: varImpPlot(cq_rf_simple)
randomForest:: varImpPlot(aa_rf_simple)
```
#________________
# below this line is future stuff, so ignore
#__________________

### Add in environmental variables here
```{r}
# commence with the modeling

  # lagged two weeks
# weekly average humidity
# weekly average temp
# weekly maximum temp
# weekly minimum temp
# weekly average precipitation
# NDVI
# NMDWI and check correlation with NDVI


# Define study extent based on occurrence data
# Add buffer around points
lon_range <- range(mosquito_final$Longitude)
lat_range <- range(mosquito_final$Latitude)
buffer <- 2  # degrees

study_extent <- extent(lon_range[1] - buffer, lon_range[2] + buffer,
                      lat_range[1] - buffer, lat_range[2] + buffer)

# Crop climate data to study area

# 
bioclim_crop <- crop(bioclim_data, study_extent)


print(names(bioclim_crop))
######
```

```{r}
## EXERCISE 4: Preparing Training Data 

# Convert occurrence coordinates to spatial points
coords_sp <- SpatialPoints(coords_filtered, 
                           proj4string = CRS("+proj=longlat +datum=WGS84"))

# Convert SpatialPoints to terra's SpatVector
coords_vect <- vect(coords_sp)

# Extract environmental values at presence locations
presence_env <- extract(env_vars, coords_vect)

# Remove any presences with missing environmental data
presence_complete <- coords_filtered[complete.cases(presence_env), ]
presence_env <- presence_env[complete.cases(presence_env), ]

nrow(presence_complete) # FInal number of presence points

# Generate pseudo-absence (background) points
# We'll use 10x the number of presences (common practice)


n_background <- (sum(mosquito_final$Presence==0) + 31203) * 10 # 10x the number of presences

# Generate random background points using terra
background_points <- spatSample(env_vars, size = n_background, 
                                method = "random", na.rm = TRUE, 
                                xy = TRUE, as.points = FALSE)

# This returns a data frame with coordinates and environmental values
# If you only want coordinates:
background_coords <- background_points[, c("x", "y")]

# Sample background points from study area
background_env <- extract(env_vars, background_coords)

# Check how many NA cells you have
sum(is.na(values(env_vars[[1]])))

# Create data frame for modeling
# Add presence/background label
presence_df <- data.frame(
  lon = presence_complete$lon,
  lat = presence_complete$lat,
  presence = 1,
  presence_env
)

background_df <- data.frame(
  lon = background_points[, 1],
  lat = background_points[, 2],
  presence = 0,
  background_env
)

# Combine presence and background data
model_data <- rbind(presence_df, background_df)

# Split data for training and testing (80/20 split)
train_indices <- sample(1:nrow(mosquito_final), size = 0.8 * nrow(mosquito_final))
train_data <- mosquito_final[train_indices, ]
test_data <- mosquito_final[-train_indices, ]

nrow(train_data)
nrow(test_data)

```

### GLM Model
```{r}
# Using all environmental predictors plus interactions
glm_model <- glm(Presence ~ tavg1 + tmin1 + 
                   tmax1 + prec1 + humd1 + 
                   NDVI + MNDWI,  # Interaction term
                 data = mosquito_final,
                 family = binomial(link = "logit"))


print(summary(glm_model))


# Predict across study area
glm_pred <- predict(env_vars, glm_model, type = "response")

# Visualize predictions
plot(glm_pred, main = "GLM: Predicted Probability of Presence",
     xlab = "Longitude", ylab = "Latitude")
points(presence_coords, pch = 20, cex = 0.3, col = "red")
map("world", add = TRUE)

```

### RF Model
```{r}
# Random Forest is a machine learning ensemble method

# Prepare data - Random Forest needs factor for classification
mosquito_final$presence_factor <- as.factor(mosquito_final$Presence)

# Build Random Forest model
# ntree = number of trees, mtry = variables per split
rf_model <- randomForest(
  Presence ~ tavg1 + tmin1 + 
    tmax1 + prec1 + 
    humd1 + NDVI + MNDWI,
  data = train_data,
  ntree = 500,
  importance = TRUE
)


print(rf_model)

# Variable importance
print(importance(rf_model))

# Plot variable importance -- mean decrease accuracy is the one that people use most
varImpPlot(rf_model, main = "Random Forest: Variable Importance")

# Predict across study area
# Need to predict probability for class "1" (presence)
rf_pred <- predict(env_vars, rf_model, type = "prob", index = 2)

# Visualize predictions
plot(rf_pred, main = "Random Forest: Predicted Probability",
     xlab = "Longitude", ylab = "Latitude")
points(presence_coords, pch = 20, cex = 0.5, col = "red")
map("world", add = TRUE)


```

### MaxEnt
```{r}

# MaxEnt is the most popular presence-only method
# Requires maxent.jar file - we'll try to run it

# Check if MaxEnt is available
maxent_path <- system.file("java", package="dismo")
maxent_path

install.packages("rJava")
library(rJava)

if (file.exists(paste0(maxent_path, "/maxent.jar"))) {
  
  # Convert presence coordinates to a matrix or data frame with just lon/lat
   presence_matrix <- as.data.frame(presence_coords[, c("lon", "lat")])
   
   # Convert SpatRaster to RasterStack (dismo requires raster package objects)
   library(raster)
   env_vars_raster <- stack(env_vars)
 
  # Build MaxEnt model
  maxent_model <- maxent(env_vars_raster, presence_matrix)
  
  print(maxent_model)
  
  # Predict across study area
  maxent_pred <- predict(maxent_model, env_vars_raster)
  
  # Visualize predictions
  plot(maxent_pred, main = "MaxEnt: Habitat Suitability",
       xlab = "Longitude", ylab = "Latitude")
  points(presence_matrix, pch = 20, cex = 0.3, col = "red")
  map("world", add = TRUE)
  
  # Plot response curves
  response(maxent_model)
  
} else {
  cat("MaxEnt not available (requires maxent.jar)\n")
  cat("To use MaxEnt:\n")
  cat("1. Download maxent.jar from: https://biodiversityinformatics.amnh.org/open_source/maxent/\n")
  cat("2. Place it in:", maxent_path, "\n")
  cat("3. Restart R\n\n")
  cat("Skipping MaxEnt for this exercise...\n")
  maxent_pred <- NULL
}

```

### Ensemble Modeling Tenatively
```{r}
## EXERCISE 11: Ensemble Modeling

# Combine multiple models using weighted average based on AUC

#. THROWS ERROR, WILL NEED ADJUSTMENT TO FIX

# Stack predictions
if (!is.null(maxent_pred)) {
  pred_stack <- stack(bioclim_pred, glm_pred, rf_pred, maxent_pred)
  weights <- auc_results$AUC / sum(auc_results$AUC)
} else {
  pred_stack <- stack(bioclim_pred, glm_pred, rf_pred)
  weights <- auc_results$AUC[1:3] / sum(auc_results$AUC[1:3])
}

# Calculate weighted mean
ensemble_pred <- weighted.mean(pred_stack, w = weights)

# Visualize ensemble prediction
plot(ensemble_pred, main = "Ensemble Model: Weighted Average",
     xlab = "Longitude", ylab = "Latitude")
points(presence_coords, pch = 20, cex = 0.3, col = "black")
map("world", add = TRUE)

# Calculate ensemble AUC
ensemble_test_pred <- extract(ensemble_pred, test_coords)
ensemble_auc <- calculate_auc(test_presence, ensemble_test_pred)
round(ensemble_auc, 3)


```

### Model Evaluation via AUC
```{r}
## EXERCISE 9: Model Evaluation

# Extract test data coordinates and environmental values
test_coords <- test_data[, c("lon", "lat")]
test_presence <- test_data$presence

# Get predictions for test data from each model

# GLM predictions
glm_test_pred <- extract(glm_pred, test_coords)

# Random Forest predictions
rf_test_pred <- extract(rf_pred, test_coords)

# MaxEnt predictions (if available)
if (!is.null(maxent_pred)) {
  maxent_test_pred <- extract(maxent_pred, test_coords)
} else {
  maxent_test_pred <- NULL
}

################################################################################
# Calculate AUC (Area Under ROC Curve)
################################################################################

# Function to calculate AUC
calculate_auc <- function(obs, pred) {
  # Remove NAs
  valid <- !is.na(pred)
  obs <- obs[valid]
  pred <- pred[valid]
  
  # Use dismo's evaluate function
  eval <- evaluate(p = pred[obs == 1], a = pred[obs == 0])
  return(eval@auc)
}

# Calculate AUC for each model
auc_results <- data.frame(
  Model = character(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)


# GLM AUC
glm_auc <- calculate_auc(test_presence, glm_test_pred)
auc_results <- rbind(auc_results, 
                     data.frame(Model = "GLM", AUC = glm_auc))

# Random Forest AUC
rf_auc <- calculate_auc(test_presence, rf_test_pred)
auc_results <- rbind(auc_results, 
                     data.frame(Model = "Random Forest", AUC = rf_auc))

# MaxEnt AUC (if available)
if (!is.null(maxent_test_pred)) {
  maxent_auc <- calculate_auc(test_presence, maxent_test_pred)
  auc_results <- rbind(auc_results, 
                       data.frame(Model = "MaxEnt", AUC = maxent_auc))
}

# Display results
# AUC Scores (Area Under ROC Curve)
# (0.5 = random, 0.7 = acceptable, 0.8 = good, 0.9 = excellent)
print(auc_results)


# Visualize AUC comparison
barplot(auc_results$AUC, 
        names.arg = auc_results$Model,
        ylim = c(0, 1),
        main = "Model Performance Comparison (AUC)",
        ylab = "AUC Score",
        col = "steelblue",
        las = 2)
abline(h = 0.7, col = "orange", lty = 2, lwd = 2)
abline(h = 0.8, col = "darkgreen", lty = 2, lwd = 2)
legend("bottomright", 
       legend = c("Acceptable (0.7)", "Good (0.8)"),
       col = c("orange", "darkgreen"),
       lty = 2, lwd = 2)

```